Vibronic coupling in C$_2$H: implications for astrochemistry

### 2022-03-31 Aug2021-CH2-.tex
update references

\end{abstract}
\section{Introduction}
Carbon monohydrides C$_{2n}$H, are a class of linear radicals that play an important role in combustion and interstellar chemistry. These carbon chains have been observed in many interstellar environments, including planetary atmospheres, comets, dark clouds, and during heavy star formation. Due to their relatively high abundance in a variety of astronomical conditions, they are believed to be promising candidates for Diffuse Interstellar Band (DIB) carriers. The corresponding anions C$_{2n}$H$^-$ are also believed to play an important role in interstellar chemistry. C$_6$H$^-$ was the first negative ion detected in space, after it was observed in two distinct interstellar molecular clouds IRC+10216 and TMC-1 in 2006. More recently, anions C$_4$H$^-$ and C$_8$H$^-$ have also been detected in a range of environments, include dark clouds, prestellar cores, and prostellar envelopes.

Chain growth of the C$_{2n}$H radicals occurs predominately through acetylene addition,
\begin{equation}
\text{C}_{2n}\text{H} + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}+\text{H}_2.
\end{equation}
Conversely, the formation of anions in space is believed to be driven by radiative electron attachment, charge transfer, and dissociate electron attachment. Current astrochemical modelling of interstellar clouds, such as IRC+1026, typically underestimate the observed abundances of CH$_4^-$. Therefore, it has been proposed that ion-neutral reactions
 \begin{equation}
	\text{C}_{2n}\text{H}^- + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}^- + \text{H}_2.
\end{equation}
may be significant chain growth mechanisms that also needs to be included. Modelling of the extraterrestrial planetary atmosphere of Saturn's moon Titan, suggests that both C$_2$H and C$_2$H$^-$ play an important role in the atmospheric chemistry. Multiple reaction pathways were included for the ion C$_2$H$^-$ including reactions with C$_2$H$_2$ and HCN.

Understanding interstellar chemistry relies critically on theory, to provide a link between astronomical observations and terrestrial laboratory studies. Microwave spectroscopy has successfully been used in this fashion to identify a large number of molecules in the interstellar medium (ISM). However for some species, particularly those with low abundances or no dipole moment, UV/vis spectroscopic methods are required for identification. This creates a challenge for theory, as electronic spectra calculations are more sensitive to the level of theory employed than pure rotational spectra. This becomes even more challenging when vibronic coupling effects are introduced, as they can be exceedingly difficult to model.

These considerations may be explored by examining some of the smallest carbon monohydrides. The ethynyl radical C$_2$H may appear to be a simple linear triatomic molecule. However the electronic spectrum is complicated by the presence of the close-lying ground $^2\Sigma^+$ and first excited $^2\Pi$ surfaces, which are only separated by $\sim3,600~$cm$^{-1}$. The interaction of these surfaces produces a complex vibronic spectrum around the $\tilde{A}^2\Pi$ state, where the $\tilde{X}-\tilde{A}$ origin is spread over multiple admixed vibronic levels. In C$_4$H the $^2\Sigma^+$ and $^2\Pi$ states are nearly degenerate, resulting in even stronger coupling, while in C$_6$H and C$_8$H the ordering of the states swaps, with a ground $^2\Pi$ state and a low lying excited $^2\Sigma^+$ state. Consequently, understanding these vibronic coupling interactions between the $^2\Sigma$ and $^2\Pi$ surfaces will be essential in order to accurately model the role these radicals (and their corresponding anions) are likely to play in the interstellar chemistry mentioned above, and to guide the search for possible DIB carriers.

In this paper we study the vibronic spectrum of the ethynyl radical C$_2$H by employing High-Resolution Photoelectron Imaging (HR-PEI) in order to benchmark CFOUR \emph{ab-initio} vibronic-coupling calculations. C$_2$H is reported to be one of the most abundant molecules in the universe, and is the most thoroughly studied of the C$_{2n}$H species. Many different experimental techniques have been employed to try and resolve the complex vibronic spectrum, including electron spin resonance, laser magnetic resonance, microwave and milli-meter wave spectroscopy, infrared (matrix isolation and Fourier Transform) spectroscopy, and photoelectron spectroscopy. C$_2$H has also received extensive theoretical attention to try and understand the experimental results, however the large number of possible vibronic levels means this has proven to be a challenge. In this work we demonstrate how construction of a quasiadiabatic Hamiltonian allows for the strength of vibronic interactions between coupled surfaces near a conical intersection to be calculated, in order to simulate electronic and vibronic spectra. By employing anion HR-PEI both the $^2\Sigma^+$ and $^2\Pi$ surfaces are mapped out on an equal footing from the anion $\tilde{X}^1\Sigma^+$ state, allowing for direct comparison to the \emph{ab-initio} modelling.

>>>>

\end{abstract}
\section{Introduction}
Carbon monohydrides C$_{2n}$H, are a class of linear radicals that play an important role in combustion and interstellar chemistry~\cite{fre85,bou96,wil91,vui01}. These carbon chains have been observed in many interstellar environments, including planetary atmospheres~\cite{vui01,wil03,dob16}, comets~\cite{jac96}, dark clouds~\cite{ziu82,wil91,gup09}, and during heavy star formation~\cite{beu08,jia15}. Due to their relatively high abundance in a variety of astronomical conditions, they are believed to be promising candidates for Diffuse Interstellar Band (DIB) carriers~\cite{dou97,ful93,wat94,ful00,sch05,gup09}. The corresponding anions C$_{2n}$H$^-$ are also believed to play an important role in interstellar chemistry~\cite{mil17,gup09}. C$_6$H$^-$ was the first negative ion detected in space, after it was observed in two distinct interstellar molecular clouds IRC+10216 and TMC-1 in 2006~\cite{mcc06}. More recently, anions C$_4$H$^-$~\cite{cer07} and C$_8$H$^-$~\cite{bru07,rem07} have also been detected in a range of environments, include dark clouds~\cite{cor13}, prestellar cores,~\cite{sak10} and prostellar envelopes~\cite{sak07}.

Chain growth of the C$_{2n}$H radicals occurs predominately through acetylene addition,
\begin{equation}
\text{C}_{2n}\text{H} + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}+\text{H}_2.
\end{equation}
Conversely, the formation of anions in space is believed to be driven by radiative electron attachment, charge transfer, and dissociate electron attachment. Current astrochemical modelling of interstellar clouds, such as IRC+1026, typically underestimate the observed abundances of CH$_4^-$~\cite{mil17,cor13,her08}. Therefore, it has been proposed that ion-neutral reactions
 \begin{equation}
	\text{C}_{2n}\text{H}^- + \text{C}_2\text{H}_2 \rightarrow \text{C}_{2n+2}\text{H}^- + \text{H}_2.
\end{equation}
may be significant chain growth mechanisms that also needs to be included~\cite{mil17,bas19}. Modelling of the extraterrestrial planetary atmosphere of Saturn's moon Titan, suggests that both C$_2$H and C$_2$H$^-$ play an important role in the atmospheric chemistry~\cite{dob16,vui09,des17,vri18}. Multiple reaction pathways were included for the ion C$_2$H$^-$ including reactions with C$_2$H$_2$ and HCN.

Understanding interstellar chemistry relies critically on theory, to provide a link between astronomical observations and terrestrial laboratory studies. Microwave spectroscopy has successfully been used in this fashion to identify a large number of molecules in the interstellar medium (ISM). However for some species, particularly those with low abundances or no dipole moment, UV/vis spectroscopic methods are required for identification~\cite{mai97}. This creates a challenge for theory, as electronic spectra calculations are more sensitive to the level of theory employed than pure rotational spectra~\cite{for10}. This becomes even more challenging when vibronic coupling effects are introduced, as they can be exceedingly difficult to model.

These considerations may be explored by examining some of the smallest carbon monohydrides. The ethynyl radical C$_2$H may appear to be a simple linear triatomic molecule. However the electronic spectrum is complicated by the presence of the close-lying ground $^2\Sigma^+$ and first excited $^2\Pi$ surfaces, which are only separated by $\sim3,700~$cm$^{-1}$~\cite{cur85,tar03}. The interaction of these surfaces produces a complex vibronic spectrum around the $\tilde{A}^2\Pi$ state, where the $\tilde{X}-\tilde{A}$ origin is spread over multiple admixed vibronic levels. In C$_4$H the $^2\Sigma^+$ and $^2\Pi$ states are nearly degenerate, resulting in even stronger coupling~\cite{zho07}, while in C$_6$H and C$_8$H the ordering of the states swaps, with a ground $^2\Pi$ state and a low lying excited $^2\Sigma^+$ state~\cite{lin99,tay98}. Consequently, understanding these vibronic coupling interactions between the $^2\Sigma$ and $^2\Pi$ surfaces will be essential in order to accurately model the role these radicals (and their corresponding anions) are likely to play in the interstellar chemistry mentioned above, and to guide the search for possible DIB carriers.

In this paper we study the vibronic spectrum of the ethynyl radical C$_2$H by employing High-Resolution Photoelectron Imaging (HR-PEI) in order to benchmark CFOUR \emph{ab-initio} vibronic-coupling calculations. C$_2$H is reported to be one of the most abundant molecules in the universe, and is the most thoroughly studied of the C$_{2n}$H species~\cite{wil91,hei99,for10}. Many different experimental techniques have been employed to try and resolve the complex vibronic spectrum, including electron spin resonance~\cite{coc64,jin85}, laser magnetic resonance~\cite{bro88,pfe96,sch98}, microwave and milli-meter wave spectroscopy~\cite{sas81,got83,end89,mul00}, infrared (matrix isolation~\cite{she87,jac87,for95} and Fourier Transform emission~\cite{ver88}) spectroscopy, laser induced fluorescence spectroscopy~\cite{hsu93,hsu95,chi99} and photoelectron spectroscopy~\cite{erv91,tay98,zho07}. C$_2$H has also received extensive theoretical attention to try and understand the experimental results, however the large number of possible vibronic levels means this has proven to be a challenge~\cite{per90,per91,per91b,per91c,per92,car00,tar03.tar04,for10,sta21}. In this work we demonstrate how construction of a quasiadiabatic Hamiltonian allows for the strength of vibronic interactions between coupled surfaces near a conical intersection to be calculated, in order to simulate electronic and vibronic spectra. By employing anion HR-PEI both the $^2\Sigma^+$ and $^2\Pi$ surfaces are mapped out on an equal footing from the anion $\tilde{X}^1\Sigma^+$ state, allowing for direct comparison to the \emph{ab-initio} modelling.
